Binning Strategies and Related Loss for Large Data

### 2015-04-08 Paper.Rnw
Frequency Binning edits: -before (1 to 7) now (1 to 10) bins -freq loss lineplot rebuild -add quantile binning of log counts -rewrite several paragraphs in section 5.2 -spell check

All rectangular binning algorithms require the specification of a bin width for each dimension and a binning origin. While using smaller bin widths leads to smaller spatial losses, reducing bin sizes comes at the cost of computation time; a potentially non-negligible consideration in settings with truly massive data. 
\ktm{ Figure~\ref{fig:SpatialLossCompare} also shows that the computation time needed to bin the simulated data sets is a decreasing function of the bin size and that random binning is marginally slower than standard binning across all bin sizes. Computation times for binning were collected using a commercial Asus laptop with an Intel Core i7 processor running at 2.80 GHz. Another pertinent arguement for using larger bins is that it will smooth over larger larger regions if visualizing the large scale density structure is of primary interest.}

@@

\subsection{Frequency Binning Specifications and Frequency Loss}
\ktm{
The reduced binned data from spatial binning contains the center and count information for all bins.  The bin frequencies may themselves be binned in order to discretize the color scale for the binned scatterplot. The loss of frequency information in frequency binning is dependent on the selection of a discrete color mapping, where the binning algorithm and number of frequency bins must be specified. To explore frequency loss characteristics using the simulated bivariate data sets, each of the three fine resolution sets were first spatially binned using standard rectangular binning to produce three sets of reduced binned data; then standard, quantile and log frequency binning algorithms were applied with various numbers of frequency bins. As discussed in Section~\ref{FreqLoss}, we will limit the considered options to using between one and seven frequency bins due to the upper limits of human perceptual abilities \citep{colorperception}.

Figure~\ref{fig:FreqLossCompare} displays the frequency loss from using standard and quantile frequency binning algorithms, with between one and seven frequency bins, from each set of simulated data. We first notice that large difference in the magnitude of frequency losses based on the bivariate distribution; frequency losses are highest for the exponential data, lower for the normal data and lowest for the uniform data. We also note that the frequency loss is a decreasing function of the number of frequency bins for both binning algorithms, thus we should consider using the maximum number of bin shades that human perception is able to simultaneously discern.  Lastly, the frequency loss with the standard frequency binning may be higher than for the quantile-based algorithm when a minimal number of bins are used; but in all three simulated data sets, this is reversed when seven frequency bins are used. Thus we recommend using the standard algorithm over the quantile-based algorithm for frequency binning because it had smaller frequency loss at the recommended seven bins, and additionally carries the benefit that equally dividing the bin count range into equally spaced bins creates a simpler shading interpretation than dividing by quantiles.

Due to the difference in scales between counts and log counts, the frequency loss and log frequency loss can not be directly compared. Frequency binning using the log counts did however behave very similarly, where log frequency loss decreased as the number of frequency bins increased.  Thus the same advise to use seven frequency bins holds for shading based on log counts as well. 
}

@@

<<FreqLossCompare,echo=F,include=T,eval=T,fig.width=10, fig.height=4, out.width='.99\\linewidth', fig.pos='H',fig.align='center',tidy=F, cache=TRUE, warning=F, fig.show='hold',fig.cap="Lineplots for total frequency loss from standard and quantile frequency binning of reduced binned data from each bivariate distribution. Spatial binning to produce the reduced binned data done with standard rectangular binning with 5 X 5 unit bins and origin at (0,0).">>= 
### Loop over ncolors 1 to 7 and also over data distribution
# find freq loss from standard, quantile, logCount
freqLossSimDat <- NULL
distrib <- c("Uniform", "Normal", "Exponential")
for (distidx in 1:3){
  binout1 <- RectBin2d(datlist[[distidx]]$x,datlist[[distidx]]$y,0,0,5,5,type="standard")
  for (ncol in 1:7){
    binoutFreq <- rbind(freqBin(binout1, binType="standard", ncolor=ncol, logCount=FALSE)[[2]],
                        freqBin(binout1, binType="quantile", ncolor=ncol, logCount=FALSE)[[2]], 
                        freqBin(binout1, binType="standard", ncolor=ncol, logCount=TRUE)[[2]])
    binoutFreq$FreqBinType <- c("Standard", "Quantile", "Log Count")
    binoutFreq$DistType <- distrib[distidx]
    binoutFreq$ncolor <- ncol 
    freqLossSimDat <- rbind(freqLossSimDat,binoutFreq)
  }
}

freqLossSimDat$FreqBinType <- factor(freqLossSimDat$FreqBinType , levels=c("Standard", "Quantile", "Log Count"))
plotdat <- freqLossSimDat[(freqLossSimDat$FreqBinType!="Log Count" & freqLossSimDat$DistType=="Exponential"),]
p1 <- qplot(ncolor, totalFreqLoss, geom="path",linetype=FreqBinType,
      data=plotdat , color = FreqBinType, size=I(1)) +
  facet_grid(.~DistType, scales="free_y") +  theme_bw() +
  scale_x_continuous(breaks=1:7) + ylim(0,max(plotdat$totalFreqLoss)) +
  xlab("Freq Bins") + ylab("Total Freq Loss") + 
  theme(legend.position="bottom")
plotdat <- freqLossSimDat[(freqLossSimDat$FreqBinType!="Log Count" & freqLossSimDat$DistType=="Normal"),]
p2 <-  qplot(ncolor, totalFreqLoss, geom="path",linetype=FreqBinType,
      data=plotdat , color = FreqBinType, size=I(1)) +
  facet_grid(.~DistType, scales="free_y") +  theme_bw() +
  scale_x_continuous(breaks=1:7) + ylim(0,max(plotdat$totalFreqLoss)) +
  xlab("Freq Bins") + ylab("Total Freq Loss") + 
  theme(legend.position="bottom")
plotdat <- freqLossSimDat[(freqLossSimDat$FreqBinType!="Log Count" & freqLossSimDat$DistType=="Uniform"),]
p3 <-qplot(ncolor, totalFreqLoss, geom="path",linetype=FreqBinType,
      data=plotdat , color = FreqBinType, size=I(1)) +
  facet_grid(.~DistType, scales="free_y") +  theme_bw() +
  scale_x_continuous(breaks=1:7) + ylim(0,max(plotdat$totalFreqLoss)) +
  xlab("Freq Bins") + ylab("Total Freq Loss") + 
  theme(legend.position="bottom")

grid.arrange(p1,p2,p3, nrow=1)
@

<<FreqLossCompare2,echo=F,include=T,eval=T,fig.width=10, fig.height=4, out.width='.99\\linewidth', fig.pos='H',fig.align='center',tidy=F, cache=TRUE, warning=F, fig.show='hold',fig.cap="Lineplots for total log frequency loss from log frequency binning of reduced binned data from each bivariate distribution. I don't think this plot will make the final cut, but it is interesting that after logging the bin counts, the log freq loss for the uniform data does not improve with more bins (might as well use one shade for all bins)">>= 
### Loop over ncolors 1 to 7 and also over data distribution
# find freq loss from standard, quantile, logCount
freqLossSimDat <- NULL
distrib <- c("Uniform", "Normal", "Exponential")
for (distidx in 1:3){
  binout1 <- RectBin2d(datlist[[distidx]]$x,datlist[[distidx]]$y,0,0,5,5,type="standard")
  for (ncol in 1:7){
    binoutFreq <- rbind(freqBin(binout1, binType="standard", ncolor=ncol, logCount=FALSE)[[2]],
                        freqBin(binout1, binType="quantile", ncolor=ncol, logCount=FALSE)[[2]], 
                        freqBin(binout1, binType="standard", ncolor=ncol, logCount=TRUE)[[2]])
    binoutFreq$FreqBinType <- c("Standard", "Quantile", "Log Count")
    binoutFreq$DistType <- distrib[distidx]
    binoutFreq$ncolor <- ncol 
    freqLossSimDat <- rbind(freqLossSimDat,binoutFreq)
  }
}

freqLossSimDat$FreqBinType <- factor(freqLossSimDat$FreqBinType , levels=c("Standard", "Quantile", "Log Count"))
plotdat <- freqLossSimDat[(freqLossSimDat$FreqBinType=="Log Count" & freqLossSimDat$DistType=="Exponential"),]
p1 <- qplot(ncolor, totalFreqLoss, geom="path",linetype=FreqBinType,
      data=plotdat , color = I(3), size=I(1)) +
  facet_grid(.~DistType, scales="free_y") +  theme_bw() +
  scale_x_continuous(breaks=1:7) + ylim(0,max(plotdat$totalFreqLoss)) +
  xlab("Freq Bins") + ylab("Total Log Freq Loss") + 
  theme(legend.position="bottom")
plotdat <- freqLossSimDat[(freqLossSimDat$FreqBinType=="Log Count" & freqLossSimDat$DistType=="Normal"),]
p2 <-  qplot(ncolor, totalFreqLoss, geom="path",linetype=FreqBinType,
      data=plotdat , color = I(3), size=I(1)) +
  facet_grid(.~DistType, scales="free_y") +  theme_bw() +
  scale_x_continuous(breaks=1:7) + ylim(0,max(plotdat$totalFreqLoss)) +
  xlab("Freq Bins") + ylab("Total Log Freq Loss") + 
  theme(legend.position="bottom")
plotdat <- freqLossSimDat[(freqLossSimDat$FreqBinType=="Log Count" & freqLossSimDat$DistType=="Uniform"),]
p3 <-qplot(ncolor, totalFreqLoss, geom="path",linetype=FreqBinType,
      data=plotdat , color = I(3), size=I(1)) +
  facet_grid(.~DistType, scales="free_y") +  theme_bw() +
  scale_x_continuous(breaks=1:7) + ylim(0,max(plotdat$totalFreqLoss)) +
  xlab("Freq Bins") + ylab("Total Log Freq Loss") + 
  theme(legend.position="bottom")

grid.arrange(p1,p2,p3, nrow=1)
@

@@

If we elect to use frequency binning to aid the perceptual mapping tile shades to the plot key for frequency, it is recommended to use seven distinct shades.  This is done to minimize the frequency loss within the bounds of human perceptual ability to distinguish multiple shades simultaneously. Standard frequency binning is recommended over the quantile frequency binning because it displayed lower frequency loss for all three simulated binned data sets while using seven frequency bins.  Log frequency binning is an alternative if the bin counts are heavily right skewed and it is desired to differentiate bins with low density.

>>>>

All rectangular binning algorithms require the specification of a bin width for each dimension and a binning origin. While using smaller bin widths leads to smaller spatial losses, reducing bin sizes comes at the cost of computation time; a potentially non-negligible consideration in settings with truly massive data. 
\ktm{ Figure~\ref{fig:SpatialLossCompare} also shows that the computation time needed to bin the simulated data sets is a decreasing function of the bin size and that random binning is marginally slower than standard binning across all bin sizes. Computation times for binning were collected using a commercial Asus laptop with an Intel Core i7 processor running at 2.80 GHz. Another pertinent argument for using larger bins is that it will smooth over larger larger regions if visualizing the large scale density structure is of primary interest.}

@@

\subsection{Frequency Binning Specifications and Frequency Loss}
\ktm{
The reduced binned data from spatial binning contains the center and count information for all bins.  The bin frequencies may themselves be binned in order to discretize the color scale for the binned scatterplot. The loss of frequency information in frequency binning is dependent on the selection of a discrete color mapping, where the binning algorithm and number of frequency bins must be specified. To explore frequency loss characteristics using the simulated bivariate data sets, each of the three fine resolution sets were first spatially binned with standard rectangular binning with 5 X 5 unit bins and origin at (0,0). Then standard and quantile frequency binning algorithms were applied to either the bin counts or log bin counts.  As discussed in Section~\ref{FreqLoss}, the average person is able to distinguish about seven colors simultaneously, so we will limit the considered options to using between one and ten frequency bins, near the upper limit of human perceptual abilities\citep{colorperception}.

The top row in Figure~\ref{fig:FreqLossCompare} displays the frequency loss from using standard and quantile frequency binning algorithms, with between one and ten frequency bins, from each set of simulated data. We first notice that large difference in the magnitude of frequency losses based on the bivariate distribution; frequency losses are highest for the exponential data, lower for the normal data and lowest for the uniform data. We also note that the frequency loss is a decreasing function of the number of frequency bins for both binning algorithms. This relationship largely flattens out after four frequency bin, each subsequent bin reducing the frequency loss less than the previous. Thus we should consider using between four and seven bin shades in order to reduce loss while also allowing for easy perception of frequency groups in the binned scatterplot. Lastly, the frequency loss with the standard frequency binning is be higher than for the quantile-based algorithm when a minimal number of bins are used. But in all three simulated data sets, the losses drop to comparable levels for each algorithm when four or more frequency bins are used.

Due to the difference in scales between counts and log counts, the frequency loss and log frequency loss can not be directly compared. The bottom row of Figure~\ref{fig:FreqLossCompare} displays the log frequency loss from using standard and quantile algorithms for binning the \textit{log} bin counts from the same sets of simulated data. Frequency binning using the log counts behaved very similarly, where log frequency loss decreased as the number of frequency bins increased. The primary difference uniform simulated data with bin counts that have low variability, are rendered virtually indistinguishable after a log transformation; thus changing the number of frequency bins does little to impact the log frequency loss. The same advise to use between four and seven frequency bins also holds for shading a binned scatterplot based on log counts.  
}

@@

<<FreqLossCompare,echo=F,include=T,eval=T,fig.width=12, fig.height=8, out.width='.99\\linewidth', fig.pos='H',fig.align='center',tidy=F, cache=TRUE, warning=F, fig.show='hold',fig.cap="Lineplots for total frequency loss (top row) and total log frequency loss (bottom row) from standard and quantile binning of bin counts and log bin counts, respectively, using reduced binned data from each bivariate distribution. ">>= 
### Loop over ncolors 1 to 10 and also over data distribution
# find freq loss from standard, quantile, logCount
freqLossSimDat <- NULL
distrib <- c("Uniform", "Normal", "Exponential")
for (distidx in 1:3){
  binout1 <- RectBin2d(datlist[[distidx]]$x,datlist[[distidx]]$y,0,0,5,5,type="standard")
  for (ncol in 1:10){
    binoutFreq <- rbind(freqBin(binout1, binType="standard", ncolor=ncol, logCount=FALSE)[[2]],
                        freqBin(binout1, binType="quantile", ncolor=ncol, logCount=FALSE)[[2]], 
                        freqBin(binout1, binType="standard", ncolor=ncol, logCount=TRUE)[[2]], 
                        freqBin(binout1, binType="quantile", ncolor=ncol, logCount=TRUE)[[2]])
    binoutFreq$FreqBinType <- c("Standard", "Quantile", "Standard Log", "Quantile Log")
    binoutFreq$DistType <- distrib[distidx]
    binoutFreq$ncolor <- ncol 
    freqLossSimDat <- rbind(freqLossSimDat,binoutFreq)
  }
}

freqLossSimDat$FreqBinType <- factor(freqLossSimDat$FreqBinType , levels= c("Standard", "Quantile", "Standard Log", "Quantile Log"))

plotdat <- freqLossSimDat[(freqLossSimDat$FreqBinType!="Standard Log" & freqLossSimDat$FreqBinType!="Quantile Log" & freqLossSimDat$DistType=="Exponential"),]
p1 <- qplot(ncolor, totalFreqLoss, geom="path",linetype=FreqBinType,
      data=plotdat , color = FreqBinType, size=I(1)) +
  facet_grid(.~DistType, scales="free_y") +  theme_bw() +
  scale_x_continuous(breaks=1:10) + ylim(0,max(plotdat$totalFreqLoss)) +
  xlab("Freq Bins") + ylab("Total Freq Loss") + 
  theme(legend.position="bottom")

plotdat <- freqLossSimDat[(freqLossSimDat$FreqBinType!="Standard Log" & freqLossSimDat$FreqBinType!="Quantile Log" & freqLossSimDat$DistType=="Normal"),]
p2 <-  qplot(ncolor, totalFreqLoss, geom="path",linetype=FreqBinType,
      data=plotdat , color = FreqBinType, size=I(1)) +
  facet_grid(.~DistType, scales="free_y") +  theme_bw() +
  scale_x_continuous(breaks=1:10) + ylim(0,max(plotdat$totalFreqLoss)) +
  xlab("Freq Bins") + ylab("Total Freq Loss") + 
  theme(legend.position="bottom")

plotdat <- freqLossSimDat[(freqLossSimDat$FreqBinType!="Standard Log" & freqLossSimDat$FreqBinType!="Quantile Log" & freqLossSimDat$DistType=="Uniform"),]
p3 <-qplot(ncolor, totalFreqLoss, geom="path",linetype=FreqBinType,
      data=plotdat , color = FreqBinType, size=I(1)) +
  facet_grid(.~DistType, scales="free_y") +  theme_bw() +
  scale_x_continuous(breaks=1:10) + ylim(0,max(plotdat$totalFreqLoss)) +
  xlab("Freq Bins") + ylab("Total Freq Loss") + 
  theme(legend.position="bottom")

plotdat <- freqLossSimDat[((freqLossSimDat$FreqBinType=="Standard Log" |freqLossSimDat$FreqBinType=="Quantile Log")  & freqLossSimDat$DistType=="Exponential"),]
p4 <- qplot(ncolor, totalFreqLoss, geom="path",linetype=FreqBinType,
      data=plotdat , color=FreqBinType, size=I(1)) +
  facet_grid(.~DistType, scales="free_y") +  theme_bw() +
  scale_x_continuous(breaks=1:10) + ylim(0,max(plotdat$totalFreqLoss)) +
  xlab("Freq Bins") + ylab("Total Log Freq Loss") + 
  theme(legend.position="bottom")

plotdat <- freqLossSimDat[((freqLossSimDat$FreqBinType=="Standard Log" |freqLossSimDat$FreqBinType=="Quantile Log") & freqLossSimDat$DistType=="Normal"),]
p5 <- qplot(ncolor, totalFreqLoss, geom="path",linetype=FreqBinType,
      data=plotdat , color = FreqBinType, size=I(1)) +
  facet_grid(.~DistType, scales="free_y") +  theme_bw() +
  scale_x_continuous(breaks=1:10) + ylim(0,max(plotdat$totalFreqLoss)) +
  xlab("Freq Bins") + ylab("Total Log Freq Loss") + 
  theme(legend.position="bottom")

plotdat <- freqLossSimDat[((freqLossSimDat$FreqBinType=="Standard Log" |freqLossSimDat$FreqBinType=="Quantile Log") & freqLossSimDat$DistType=="Uniform"),]
p6 <-qplot(ncolor, totalFreqLoss, geom="path",linetype=FreqBinType,
      data=plotdat , color = FreqBinType, size=I(1)) +
  facet_grid(.~DistType, scales="free_y") +  theme_bw() +
  scale_x_continuous(breaks=1:10) + ylim(0,40) +
  xlab("Freq Bins") + ylab("Total Log Freq Loss") + 
  theme(legend.position="bottom")

grid.arrange(p1,p2,p3,p4,p5,p6, nrow=2)
@

@@

If we elect to use frequency binning to aid the perceptual mapping tile shades to the plot key for frequency, it is recommended to use between four and seven distinct shades.  This is done to minimize the frequency loss within the bounds of human perceptual ability to distinguish multiple shades simultaneously. The standard and quantile algorithms for frequency binning have very similar loss properties for this range, so the decision falls to the interpretability of the resulting frequency bins. Log frequency binning is an alternative if the bin counts are heavily right skewed and it is desired to differentiate bins with low density.
